---
title: "2_LinksandKeywordsReport"
format: html
editor: visual
---

## Packages

```{r}
#| echo: false
library(xml2)
library(stringr)
library(dplyr)
library(purrr)
library(bib2df)
```

## Configuration

Just three settings, so you can point the script at different input/output files without touching the logic below.

```{r}
# BibTeX metadata file exported from Zotero
bib_file <- "all_articles_metadata.bib"

# Folder containing GROBID TEI-XML files for the pilot articles
xml_folder <- "Articles All Records"

# Output RDS file to be read by RUN_ME.qmd
output_file <- "auto_report_input_data.rds"
```

## Read metadata from BibTeX

Loads `all_articles_metadata.bib` with `bib2df`, then keeps `title`, `abstract`, `authors`, `year`, `journal`, `DOI`. Strips stray `{}` from titles. **Drops any entry without a DOI** (`filter(!is.na(DOI) & nzchar(DOI))`). Also builds `doi_no_slash` (DOI with `/` removed), which is the join key used everywhere else in the script. `authors`/`year`/`journal` aren't used by the report itself — they're carried through to `full_data`/`auto_report_input_data.rds` purely so `generate_coder_batch.R` can pre-fill the protocol's metadata fields (via `R/batch_json.R`) when it later packages a batch for a coder, without needing to re-read the `.bib` file itself.

```{r}
#| echo: false

bib_raw <- bib2df(bib_file)

meta_data <- bib_raw %>%
  transmute(
    title    = str_remove_all(TITLE, "[{}]"),
    abstract = ABSTRACT,
    authors  = AUTHOR,  # list-column: character vector of "Last, First" per paper
    year     = YEAR,
    journal  = JOURNAL,
    DOI      = DOI
  ) %>%
  filter(!is.na(DOI) & nzchar(DOI)) %>%
  mutate(doi_no_slash = gsub("/", "", DOI))

message("Extracted ", nrow(meta_data), " items with a DOI from BibTeX")

# Reading in inclusion data to create a vector of the included doi's
included_dois <- readRDS("all_articles.rds") %>%
  filter(included == TRUE) %>%
  pull(doi_no_slash)

message("Included ", length(included_dois), " items")

# Filter the meta_data df to contain only those items that are in the list of included doi's
meta_data <- meta_data %>%
  filter(doi_no_slash %in% included_dois) %>%
  # Add ID variable to the data frame
  mutate(ID = sprintf("P%03d", row_number()))

message("Filtered meta data to include ", nrow(meta_data), " items")
```

## Parse TEI XML files

-   **Title / abstract**: pulled from the XML as a fallback in case BibTeX is missing them.

-   **Full text**: grabs all `<p>` and `<head>` elements from the body and back matter (e.g. data-availability statements), skipping tables and formulas. Crucially, wherever GROBID found a hyperlink (`<ref type="url">`), it appends `[the-actual-url]` right after the visible anchor text — so a link that displays as "see our repository" becomes "see our repository \[https://osf.io/...\]" in the plain-text version. That's what lets later keyword-context snippets show clickable links even though they're just strings by that point.

-   **URLs**: separately extracted into a `url_data` table of `{url, context}` pairs, using the parent paragraph as context — this is the primary/cleaner source of links (vs. text-mining full_text for `https://` later).

-   Wrapped in `tryCatch` — if a file fails to parse, it logs a warning and returns an all-NA row instead of crashing the whole loop.

Each file's result becomes one row of `df_fulltext`. The `doi_no_slash` for that row is always derived from the filename (GROBID names files `<original>.pdf.tei.xml`, so it strips that suffix), never from DOI text inside the document — GROBID sometimes mis-extracts a different DOI from the header (e.g. a supplementary-materials DOI) or from bibliography references, which would silently break the join. The filename is the reliable join key: files are named by `doi_no_slash` in the first place, matching `meta_data$doi_no_slash` (from BibTeX) and `all_articles.rds$doi_no_slash`.

```{r}
#| echo: false

clean_url <- function(url) str_remove_all(url, "[​‌‍­﻿]")

parse_tei_xml <- function(xml_path) {
  tryCatch({
    doc <- read_xml(xml_path)
    ns  <- c(tei = "http://www.tei-c.org/ns/1.0")

    # Title (use as fallback if not in RDF)
    title_node <- xml_find_first(
      doc, "//tei:titleStmt/tei:title[@level='a'][@type='main']", ns)
    if (is.na(title_node))
      title_node <- xml_find_first(doc, "//tei:titleStmt/tei:title", ns)
    title <- if (!is.na(title_node)) trimws(xml_text(title_node)) else NA_character_

    # Abstract (fallback)
    abs_nodes <- xml_find_all(doc, "//tei:profileDesc//tei:abstract//tei:p", ns)
    abstract  <- paste(xml_text(abs_nodes), collapse = " ")
    if (nchar(trimws(abstract)) == 0) abstract <- NA_character_

    # Append "[target-url]" after a ref's anchor text so that links surface as
    # clickable text in any downstream context extracted from full_text
    # (e.g. keyword match contexts), matching how url_data does it below.
    node_text_with_links <- function(node) {
      txt  <- trimws(xml_text(node))
      refs <- xml_find_all(node, ".//*[@type='url']", ns)
      for (ref in refs) {
        target <- clean_url(xml_attr(ref, "target"))
        anchor <- trimws(xml_text(ref))
        if (!is.na(target) && nzchar(target) && nzchar(anchor) && anchor != target)
          txt <- str_replace(txt, fixed(anchor), paste0(anchor, " [", target, "]"))
      }
      txt
    }

    # Full text from body + back matter
    body_nodes <- xml_find_all(
      doc, "//tei:body//*[self::tei:p or self::tei:head][not(ancestor::tei:figure[@type='table'])][not(ancestor::tei:formula)]", ns)
    back_nodes <- xml_find_all(
      doc, "//tei:back//*[self::tei:p or self::tei:head][not(ancestor::tei:listBibl)][not(ancestor::tei:figure[@type='table'])][not(ancestor::tei:formula)]", ns)
    full_text  <- gsub("\n", " ",
      paste(c(vapply(body_nodes, node_text_with_links, character(1)),
              vapply(back_nodes, node_text_with_links, character(1))), collapse = " "))

    # URLs: extract target attribute and parent paragraph as context
    url_refs <- xml_find_all(doc, "//*[@type='url']", ns)
    url_data <- map_dfr(url_refs, function(ref) {
      target <- clean_url(xml_attr(ref, "target"))
      if (is.na(target) || !nzchar(target)) return(NULL)
      target  <- str_remove(target, "[.,;:!?]+$")
      context <- trimws(xml_text(xml_parent(ref)))
      anchor  <- trimws(xml_text(ref))
      if (nzchar(anchor) && anchor != target)
        context <- str_replace(context, fixed(anchor),
                               paste0(anchor, " [", target, "]"))
      tibble(url = target, context = context)
    })

    all_urls <- if (nrow(url_data) > 0) unique(url_data$url) else character(0)

    list(title = title, abstract = abstract, full_text = full_text,
         urls = all_urls, url_data = url_data)

  }, error = function(e) {
    warning("Failed to parse: ", basename(xml_path), " — ", e$message)
    list(title = NA_character_, abstract = NA_character_, full_text = NA_character_,
         urls = character(0),
         url_data = tibble(url = character(0), context = character(0)))
  })
}

xml_paths <- list.files(xml_folder,
                        pattern = "\\.(tei\\.xml|xml)$",
                        recursive = TRUE,
                        full.names = TRUE)

message("Found ", length(xml_paths), " XML files in ", xml_folder)

df_fulltext <- map_dfr(xml_paths, function(xml_file) {
  result <- parse_tei_xml(xml_file)

  # GROBID appends .tei.xml to the full original filename (including .pdf),
  # so strip .pdf.tei.xml first, then fall back to .tei.xml or .xml. The
  # filename is always the join key — never DOI text parsed from inside the
  # document, which GROBID can mis-extract (e.g. picking up a bibliography
  # reference's DOI or a supplementary-materials DOI instead of the article's).
  doi_no_slash <- basename(xml_file) %>%
    str_replace("\\.pdf\\.tei\\.xml$|\\.tei\\.xml$|\\.xml$", "")

  tibble(
    doi_no_slash = doi_no_slash,
    title_xml    = result$title,
    abstract_xml = result$abstract,
    full_text    = result$full_text,
    urls         = list(result$urls),
    url_data     = list(result$url_data)
  )
})

message("Parsed ", nrow(df_fulltext), " XML files")
```

## Join metadata with full text

Left-joins the BibTeX metadata (`meta_data`) with the XML-parsed data (`df_fulltext`) on `doi_no_slash`. Title/abstract: BibTeX wins, XML fills gaps (`coalesce`). Builds a clickable `DOI Link` column. Logs how many of the BibTeX articles actually found a matching XML file — worth checking that number isn't unexpectedly low.

```{r}
full_data <- meta_data %>%
  left_join(df_fulltext, by = "doi_no_slash") %>%
  mutate(
    # BibTeX metadata takes priority; fall back to XML-extracted values
    title    = coalesce(title, title_xml),
    abstract = coalesce(abstract, abstract_xml),
    `DOI Link` = paste0("https://doi.org/", DOI)
  ) %>%
  select(-title_xml, -abstract_xml)

message(sum(!is.na(full_data$full_text)), " of ", nrow(full_data),
        " articles matched to an XML file")
```

## Keyword definitions

```{r}
key_phrases_exact <- c("code", "script", "osf", "git",
                       "zenodo", "figshare", "supplement", "replicate",
                       "reproduce", "appendix", "syntax", "upon request",
                       "restricted access", "upon reasonable request", "data availability",
                       "supporting materials", "materials", "repository")
    
key_phrases_gapped <- list(
  c("available", "request"),
  c("code", "request"),
  c("upon", "request"),
  c("supplement", "request"),
  c("availability", "statement")
)
max_gap <- 3
```

## Extract keyword context

Splits each article's full text into sentences, then for every keyword/pair, finds matching sentences and grabs a 3-sentence window (previous + matched + next) as context. Runs this over every article's `full_text` via `map()`, storing the result as a nested tibble in a new `keyword_data` list-column. Also bolts on four empty screening columns every article needs downstream: `included`, `excluded`, `second_reviewer` (all `FALSE`), `note` (`""`) — these are what you'll fill in by hand while reviewing in `RUN_ME.qmd`'s output.

```{r}
extract_keyword_context <- function(text, key_phrases_exact, key_phrases_gapped, max_gap) {
  if (is.na(text) || nchar(trimws(text)) == 0)
    return(tibble(keyword = character(0), context = character(0)))

  sentences <- unlist(strsplit(text, "(?<=[.!?])\\s+", perl = TRUE))

  # Returns one row per unique sentence that matches pattern, labelled with kw_label
  contexts_for <- function(kw_label, pattern) {
    idx <- unique(which(str_detect(sentences, pattern)))
    if (length(idx) == 0) return(tibble(keyword = character(0), context = character(0)))
    map_dfr(idx, function(i) {
      tibble(
        keyword = kw_label,
        context = paste(na.omit(c(
          if (i > 1) sentences[i - 1] else NA_character_,
          sentences[i],
          if (i < length(sentences)) sentences[i + 1] else NA_character_
        )), collapse = " ")
      )
    })
  }

  # Prefix match (\bkeyword) catches all case variants and suffixes:
  # e.g. "supplement" also matches Supplement, supplementary, supplemental
  exact_results <- map_dfr(key_phrases_exact, function(kw) {
    contexts_for(kw, regex(paste0("\\b", kw), ignore_case = TRUE))
  })

  gapped_results <- map_dfr(key_phrases_gapped, function(pair) {
    pattern <- regex(
      paste0("\\b", pair[1], "(?:\\s+[\\w,\\.\\!\\?]+){0,", max_gap, "}\\s+\\b", pair[2]),
      ignore_case = TRUE
    )
    contexts_for(paste(pair, collapse = " "), pattern)
  })

  bind_rows(exact_results, gapped_results)
}

full_data <- full_data %>%
  mutate(
    keyword_data = map(full_text, ~extract_keyword_context(
      .x, key_phrases_exact, key_phrases_gapped, max_gap)),
    included        = FALSE,
    excluded        = FALSE,
    second_reviewer = FALSE,
    note            = ""
  )

message("Keyword extraction complete")
```

## Save pilot data

Writes the whole thing to `auto_report_input_data.rds`. That file is then read straight in by `RUN_ME.qmd` to generate the report.

```{r}
saveRDS(full_data, file = output_file)
message("Saved ", nrow(full_data), " articles to: ", output_file)
```
