---
title: "1_pdf_to_XML"
format: html
editor: visual
---

> **Do not run this script yourself.** It is run by `GENERATE_BATCH.R` in the project folder, which also holds all its settings. Paths below are relative to the project folder.

Step 1 of the pipeline: converts the PDFs to TEI-XML via GROBID.

## Dependency Management

```{r}

renv::restore()

library(metacheck)
```

## Configuration

Settings come from `GENERATE_BATCH.R`. The values after `%||%` are only fallbacks for when this script is run on its own, which is not needed.

```{r}
cfg <- if (exists("batch_settings", envir = globalenv())) get("batch_settings", envir = globalenv()) else list()

# Folder with the PDFs (searched including subfolders)
pdf_folder <- cfg$pdf_folder %||% "Pilot3/files"

# Folder the XML files are written to (read by step 2)
xml_folder <- cfg$xml_folder %||% "Pilot3"

# GROBID server. If it is not active, take another one from the list of active
# servers: https://www.scienceverse.org/metacheck/convert.json
grobid_url <- cfg$grobid_url %||% "https://grobid.hti.ieis.tue.nl"
```

## Process batch to XML files

Only PDFs that do not have an XML file in `xml_folder` yet are converted, so re-running does not send every PDF to the GROBID server again. To convert a PDF again, delete its XML file first.

```{r}

#Listing all the filepaths at which the pdf's are stored
pdf_files <- list.files(
  path = pdf_folder,
  pattern = "\\.pdf$",
  recursive = TRUE,
  full.names = TRUE
)

# Skip PDFs that were already converted (<name>.pdf -> <name>.xml)
done <- file.exists(file.path(xml_folder, paste0(tools::file_path_sans_ext(basename(pdf_files)), ".xml")))
message(length(pdf_files), " PDFs found in ", pdf_folder, "; ", sum(done),
        " already converted, ", sum(!done), " to convert")
pdf_files <- pdf_files[!done]

# Processing the batch of files to XML files
if (length(pdf_files) > 0) {
  paper_testbatch <- convert(file_path = pdf_files,
                             save_path = xml_folder,
                             method = "grobid",
                             api_url = grobid_url,
                             crossref_lookup = FALSE)
}
```
